### 2018-01-09 submission/manuscript.Rmd
re write first result section

***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
By power transforming and Z-score normalizing the $\alpha$-diversity metrics for the entire data set we assessed whether there were any broad scale community differences that could be detected as disease severity worsened. Using linear mixed-effect models to control for study, re-sampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) and no $\alpha$-diversity correlations in tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1]. We next tested whether these detectable differences in community resulted in significant increases in RR. For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].

Using the Bray-Curtis distance metric, there was a significant difference across study in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S4 & S5]. For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with tissue samples from the same individual (matched) had no differences [Table S6 & S7].

>>>>

***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
Power transforming and Z-score normalizing OTU richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess for differences in these broad scale community metrics as disease severity worsens. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased. There was a significant decrease in both evenness and Shannon diversity from control to adenoma to carcinoma in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether these detectable differences in community resulted in significant increases in RR. For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma versus control and adenoma versus control [Table S1 & S2].

Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased. For tissue there were no significant changes in OTU richness, evenness, and Shannon diversity from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C]. When analyzing the RR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) tissue samples, for both adenoma and carcinoma there was no significant change in RR based on lower than median values for OTU richness, evenness, and Shannon diversity [Table S3-S5]. Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were also identified in unmatched tissue samples  for both those with adenoma and carcinoma [Table S6 & S7]. For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S6 & S7].

### 2018-01-09 submission/manuscript.tex
re write first result section

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} By power transforming and Z-score normalizing the
\(\alpha\)-diversity metrics for the entire data set we assessed whether
there were any broad scale community differences that could be detected
as disease severity worsened. Using linear mixed-effect models to
control for study, re-sampling of the same individual (tissue studies
only), and variable region, there was a significant decrease from
control to adenoma to carcinoma for both evenness and Shannon diversity
in stool (P-value = 0.025 and 0.043, respectively) and no
\(\alpha\)-diversity correlations in tissue (P-value \textgreater{}
0.05) {[}Figure 1{]}. We next tested whether these detectable
differences in community resulted in significant increases in RR. For
fecal samples, a decrease in evenness resulted in a significantly
increased RR for carcinoma (RR = 1.36 (1.15 - 1.61), P-value = 0.00034)
and adenoma (RR = 1.16 (1.01 - 1.34), P-value = 0.032) while a decrease
in Shannon diversity only increased the RR for carcinoma (RR = 1.33
(1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}. Interestingly, for both
adenoma and carcinoma there was no increase in RR within tissue samples
for any alpha diversity metric investigated {[}Table S1-S3{]}.

Using the Bray-Curtis distance metric, there was a significant
difference across study in the bacterial community of fecal samples
between carcinoma and controls, but not adenoma and controls {[}Table S4
\& S5{]}. For studies with unmatched tissue samples a similar trend was
observed {[}Table S3 \& S4{]} while studies with tissue samples from the
same individual (matched) had no differences {[}Table S6 \& S7{]}.

>>>>

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} Power transforming and Z-score normalizing OTU
richness, evenness, and Shannon diversity for the entire data set
allowed for the combination of data from all studies to assess for
differences in these broad scale community metrics as disease severity
worsens. Using linear mixed-effect models to control for study and
variable region we assessed whether OTU richness, evenness, or Shannon
diversity changed in a step-wise manner as disease severity increased.
There was a significant decrease in both evenness and Shannon diversity
from control to adenoma to carcinoma in stool (P-value = 0.025 and
0.043, respectively) {[}Figure 1A{]}. We next tested whether these
detectable differences in community resulted in significant increases in
RR. For fecal samples, a decrease in evenness resulted in a
significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
0.032) while a decrease in Shannon diversity only increased the RR for
carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
Using the Bray-Curtis distance metric and PERMANOVA, it was also
possible to identify significant bacterial community changes, in
specific studies, for both carcinoma versus control and adenoma versus
control {[}Table S1 \& S2{]}.

Using similar transformations for tissue samples, linear mixed-effect
models were used on the transformed combined data to control for study,
re-sampling of the same individual, and variable region to test whether
OTU richness, evenness, or Shannon diversity changed in a step-wise
manner as disease severity increased. For tissue there were no
significant changes in OTU richness, evenness, and Shannon diversity
from control to adenoma to carcinoma (P-value \textgreater{} 0.05)
{[}Figure 1B \& C{]}. When analyzing the RR, for matched (unaffected
tissue and an adenoma or carcinoma from the same individual) and
unmatched (control and adenoma or carcinoma tissue not from the same
individual) tissue samples, for both adenoma and carcinoma there was no
significant change in RR based on lower than median values for OTU
richness, evenness, and Shannon diversity {[}Table S3-S5{]}. Similar to
stool samples, significant differences in bacterial community, assessed
by PERMANOVA, were also identified in unmatched tissue samples for both
those with adenoma and carcinoma {[}Table S6 \& S7{]}. For studies with
matched samples no differences in bacterial community were observed when
assessed with PERMANOVA {[}Table S6 \& S7{]}.
